---
title: "births"
format: docx
editor: visual
date: now
execute: 
  echo: false
  warning: false
  cache: true
---

```{r}
library(tidyverse)
library(gt)
library(gtsummary)

births <- readxl::read_xlsx('births.xlsx',
                            na = c('','not indicated','Not Indicated',
                                   'not Indicated','Unkown','Not indicated'))
color_pal <- RColorBrewer::brewer.pal(7,'Dark2')
theme_set(theme_bw())
```

Brief overview of variables measured:

17 variables were measured across 141 patients.

```{r}
births<- births %>% 
  separate(parity,into = c('prev_term','prev_pre_term'),sep = '[+]') %>% 
  mutate(birth_weight=case_when(
    birth_weight>=2500 & birth_weight<=3000~'2500-3000g',
    birth_weight>3000~'>3000g',
    birth_weight<2500~'<2500g',
    TRUE~birth_weight
  ),
  education=str_to_title(education),
  birth_weight=str_replace(birth_weight,'g',''),
  birth_weight=factor(birth_weight,
                            levels=c('<2500','2500-3000','>3000')),
  gestational_age=as.numeric(str_replace_all(gestational_age,'weeks','')),
  across(.cols = c(occupation,education),.fns = ~str_to_title(.)),
  occupation=str_replace_all(occupation,'Self Employed','Self-Employed'),
  mode_of_delivery=if_else(mode_of_delivery=='Ceserian','Caesarean section',mode_of_delivery),
  mode_of_delivery=
    if_else(mode_of_delivery%in%c('Assisted delivery','Spontaneous Vertex Delivery'),'Spontaneous Vertex Delivery','Caesarean section'),
  nature=if_else(gestational_age>24&gestational_age<37,'Pre-term','Term'),
  across(where(is.character),as_factor)
  )
births %>% glimpse()
```

Overall age is somewhat evenly distributed

```{r}
births %>% 
  ggplot(aes(mother_age)) +
  geom_histogram(bins = 20,fill='midnightblue',color='white')+
  labs(x="Mother's age",y='Count',title = "Histogram of mothers' age")
```

```{r}
births %>% 
  ggplot(aes(mother_age)) +
  geom_histogram(bins = 20,fill='midnightblue',color='white')+
  facet_wrap(~nature)+
  labs(x="Mother's age",y='Count',title = "Histogram of mothers' age")
```

Level of education was missing for a significant number of the interviewed mothers. A few made it to university while most had completed secondary school.

```{r}
births %>% 
    select(nature,education) %>% 
    group_by(nature,education) %>% 
    summarise(n=n()) %>% 
    ggplot(aes(reorder(education,n),n))+
    geom_col(aes(fill=education),show.legend = F)+
    scale_fill_manual(values = color_pal)+
    facet_wrap(~nature)+
    labs(y='Count',x='Educational attainment')+
    coord_flip()
```

More mothers were married.

```{r}
#/ warning: false
births %>% 
  ggplot(aes(marital_status))+
  geom_bar(aes(fill=marital_status))+
  scale_fill_manual(values = color_pal)+
  labs(x='Marital Status',y='Count')
```

Among mothers who reported, most were stay at home moms, followed by the self-employed and employed.

```{r}
births %>% 
  group_by(nature,occupation) %>% 
  ggplot(aes(occupation,fill=occupation))+
  geom_bar(show.legend = F)+
  scale_fill_manual(values = color_pal)+
  coord_flip()+
  facet_wrap(~nature)+
  labs(y='Count',x='Occupation')
```

A significant number of mothers were in the mid to late thirties.

```{r}
births %>% 
  ggplot(aes(gestational_age))+
  geom_histogram(bins=10,fill='midnightblue',color='white')+
  labs(x='Gestational age',y='Frequency')
```

```{r}
births %>% 
  ggplot(aes(gestational_age))+
  geom_boxplot()+
  labs(x='Gestational Age')
```

There were more births that had complications

```{r}
births %>% 
  ggplot(aes(complication,fill=complication))+
  geom_bar(show.legend = F)+
  scale_fill_manual(values = color_pal)+
  labs(x='Complication?',y='Frequency')
```

Onset of labor for most mothers was spontaneous

```{r}
births %>% 
  select(nature,onset_of_labor) %>% 
  na.omit() %>% 
  ggplot(aes(onset_of_labor))+
  geom_bar(aes(fill=onset_of_labor),show.legend = F)+
  scale_fill_manual(values = color_pal)+
  facet_wrap(~nature)+
  labs(x='Onset of labor')
```

Complications were common among women with induced on set of labor (59%) compared to those who had spontaneous onset of labor(40%)

```{r}
births %>% 
  select(onset_of_labor,complication) %>% 
  group_by(onset_of_labor) %>% 
  count(complication) %>% 
  mutate(prop=n/sum(n)) %>% 
  ggplot(aes(reorder(onset_of_labor,-n),n))+
  geom_col(aes(fill=complication))+
  geom_text(data = . %>% filter(onset_of_labor=='Spontaneous'),
            aes(label=n),vjust=2)+
  annotate(x='Induced',y=22,geom = 'text',label=17)+
  scale_fill_manual(values = color_pal,name='Complication?')+
  labs(x='Onset of labor',y='Frequency')
```

Moreover, most of the pre-term and term births were by Caesarian Section . No pre-term births reported were by spontaneous vertex delivery.

```{r}
births %>% 
  select(nature,mode_of_delivery) %>% 
  group_by(nature,mode_of_delivery)%>%
  summarise(n=n()) %>% 
  na.omit() %>% 
  ggplot(aes(x=reorder(mode_of_delivery,n),y=n))+
  geom_col(aes(fill=mode_of_delivery),show.legend = F)+
  #coord_flip()+
  scale_fill_manual(values = color_pal)+
  scale_x_discrete(
    labels = c('Spontaneous Vertex\nDelivery','Cesarian Section'))+
  facet_wrap(~nature)+
  labs(x='Mode of Delivery')
```

A majority of deliveries were spontaneous for term and pre-term births

```{r}
births %>% 
  select(nature,onset_of_labor) %>% 
  filter(onset_of_labor!='Unknown') %>% 
  ggplot(aes(nature,fill=onset_of_labor))+
  geom_bar()+
  scale_fill_manual(values = color_pal)+
  labs(x='')
```

For the interviewed mothers, the were more male newborns than female newborns.

```{r}
births %>% 
  ggplot(aes(sex,fill=sex))+
  geom_bar()+
  scale_fill_manual(values = color_pal)+
  labs(x='',y='Count')
```

Most new borns weighed less than 2500 grams compared 17 who weighed more than 3000 grams

```{r}
births %>% 
  count(birth_weight) %>% 
  ggplot(aes(x=birth_weight,y=n,fill=birth_weight))+
  geom_col()+
  geom_text(aes(label=n),vjust=-.6)+
  scale_fill_manual(values = color_pal)+
  ylim(c(0,90))+
  labs(x='Birth Weight',y='Frequency')
```

However, differences in weight across male and female newborns are not significant

```{r}
births %>% 
  select(nature,birth_weight) %>% 
  tbl_summary(by=nature,label=list(birth_weight='Birth Weight')) %>% 
  add_p()
```

The distribution across both sexes of the newborns is almost similar for both pre-term and term births.

```{r}
births %>% 
  ggplot(aes(nature,fill=sex))+
  geom_bar()+
  scale_fill_manual(values = color_pal)+
  labs(x='',y='Count')
```

```{r}
ttt_ifas <- readxl::read_xlsx('factors.xlsx') %>% 
  select('Tetanus Toxoid','IFAS','Hb')
```

```{r}
ttt_ifas %>% 
  mutate(Hb=as.numeric(Hb))%>% 
  summarise(N=n(),
            Average=mean(Hb,na.rm=T)) %>% 
  gt() %>% 
  tab_header(title = 'Hb levels reported')
```

There is a significant association between the number of antenatal clinic visits

```{r}
births %>% 
  select(nature,anc_visits) %>% 
  mutate(anc_visits=as_factor(anc_visits)) %>% 
  filter(anc_visits!='Unknown',nature!='Unknown') %>% 
  tbl_summary(by=nature,label=list(anc_visits='Clinic Visits')) %>% 
  add_p()
```

Mother's age is not a risk factor influencing the occurrence of term and pre-term births in the sample considered.

```{r}
births %>% 
  select(nature,mother_age) %>% 
  tbl_summary(by=nature,
              label=list(mother_age="Mother's Age")) %>% 
  add_p()
```

```{r}
births %>% 
  select(nature,onset_of_labor) %>% 
  filter(onset_of_labor!='Unknown') %>% 
  tbl_summary(by=nature,label = list(onset_of_labor='Onset of Labor')) %>% 
  add_p()
```

There is close association between complications and the mode of delivery.

```{r}
births %>% 
  select(mode_of_delivery,complication) %>% 
  tbl_summary(by=complication,
              label = list(mode_of_delivery~'Mode of Delivery')) %>% 
  add_p() %>% 
  as_gt() %>% 
  tab_header(title = md('**Complication?**'))
  
```

Mode of delivery is a highly significant risk factor influencing the occurrence of term and pre-term births as confirmed by a Chi-squared test.

```{r}
births %>% 
  select(mode_of_delivery,nature) %>% 
  tbl_summary(by=nature,
              label = list(mode_of_delivery~'Mode of Delivery')) %>% 
  add_p()
```
